---
title: "Fonio, Climate, and Genes Under Selection"
author: "Amy Jackson"
date: "2024-07-29"
output: html_document
---

```{r setup, include=FALSE}
# Setting up the R environment and ensuring that code chunks are echoed
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, warning=FALSE,message=FALSE}
library(raster)
library(rgdal)
library(terra)
library(geodata)
library(predicts)
library(LEA)
library(sf)
library(terra)
library(dplyr)
library(spData)
library(tmap)
library(cartogram)
library(tidyverse)
library(gradientForest)
library(factoextra)
library(dplyr)
library(data.table)
library(readr)
library(vegan)
```

## Step 1: Data Cleaning and Preparation
### Downloading and Preparing Bioclimatic Data
We start by downloading WorldClim bioclimatic variables at a resolution of 2.5 minutes. These variables will be used to assess the environmental conditions at the locations of our species observations.

```{r bioclim part 1}
# Download WorldClim bioclimatic variables at a resolution of 2.5 minutes
# and save them in the "data/" directory
bioclim_data <- worldclim_global(var = "bio", res = 2.5, path = "data/")

# Extract the coordinate reference system (CRS) from the bioclimatic data
crs_bioclim <- crs(bioclim_data)

# Read observational data from a CSV file containing species coordinates
obs_data <- read.csv("digitaria_coords.csv")

# What does the data look like?
head(obs_data)

# Remove rows with missing latitude or longitude values
obs_data <- obs_data[!is.na(obs_data$Latitude) & !is.na(obs_data$Longitude), ]

# Convert the observational data to a simple features (sf) object
# using the latitude and longitude columns, and set the CRS to WGS 84 (EPSG:4326)
obs_sf <- st_as_sf(obs_data, coords = c("Longitude", "Latitude"), crs = 4326)

# Create a data frame with longitude and latitude columns from the observational data
coords <- data.frame(lon = obs_data$Longitude, lat = obs_data$Latitude)

# Check if points are within the extent of the raster
bioclim_extent <- ext(bioclim_data)
within_extent <- coords$lon >= bioclim_extent$xmin & coords$lon <= bioclim_extent$xmax &
  coords$lat >= bioclim_extent$ymin & coords$lat <= bioclim_extent$ymax

if (!all(within_extent)) {
  warning("Some points fall outside the extent of the raster data.")
  coords <- coords[within_extent, ]
}

# Convert filtered points back to SpatVector
points_vect <- vect(coords, geom = c("lon", "lat"), crs = crs_bioclim)
```

### Extracting Bioclimatic Values
Here, we extract bioclimatic values at the observation points, which will later be used for PCA and other analyses.

```{r bioclim part 2, echo=TRUE}
# Extract bioclimatic values at observation points
extracted_values <- terra::extract(bioclim_data, points_vect)

# Combine extracted values with observation data
obs_data_filtered <- obs_data[within_extent, ]
obs_data <- cbind(obs_data_filtered, extracted_values)

# Print the observation data with extracted bioclimatic values
head(obs_data)

obs_data <- write.csv(obs_data, "obs_data.csv")
```

## Step 2: Visualization with Maps
### Plotting Observation Data on Maps
We visualize the observation data on a map, overlaying it with environmental variables such as annual precipitation and temperature.

```{r Map plotting, echo=TRUE}

# Define the geographic extent based on observation data
max_lat <- ceiling(max(obs_data$Latitude))
min_lat <- floor(min(obs_data$Latitude))
max_lon <- ceiling(max(obs_data$Longitude))
min_lon <- floor(min(obs_data$Longitude))

geographic_extent <- ext(c(min_lon, max_lon, min_lat, max_lat))
sample_extent <- geographic_extent * 2

# Crop the climate data based on your extent.
bioclim_data <- crop(bioclim_data, sample_extent)

# Check GPS points. Load and crop world map
world_map <- world(resolution = 3, path = "data/")
my_map <- crop(world_map, sample_extent)
plot(my_map, axes = TRUE, col = "grey95")
points(x = obs_data$Longitude, y = obs_data$Latitude, col = "black", pch = 20, cex = 1)

# Check viz on your BioClim variables.
blue_palette <- colorRampPalette(c("lightblue", "blue", "darkblue"))
plot(bioclim_data[[12]], main="Annual Precipitation", col = blue_palette(100))
plot(my_map, axes = FALSE, add=TRUE)
points(x = obs_data$Longitude, y = obs_data$Latitude, pch = 20, cex = 0.75, col = "red")

red_palette <- colorRampPalette(c("blue", "lightblue", "yellow", "orange", "red"))
plot(bioclim_data[[1]], main="Annual Temperature", col = red_palette(100))
plot(my_map, axes = FALSE, add=TRUE)
points(x = obs_data$Longitude, y = obs_data$Latitude, pch = 20, cex = 0.75, col = "black")
```

## Step 3: Principal Component Analysis (PCA)
### Performing PCA on Bioclimatic Data
We conduct a PCA to explore the relationships between different bioclimatic variables and the geographic distribution of Fonio.

```{r PCA, echo=TRUE}

pca_data <- read.csv("obs_data.csv", header=TRUE)

# Summarise number of accessions we have from each Country
country_counts <- pca_data %>%
  group_by(Country) %>%
  summarize(count = n(), .groups = 'drop')

print(country_counts)

# Exploring the data
ggplot(pca_data, aes(x=Longitude, y=Latitude,colour=bio1)) +
geom_point(alpha=0.5) +
facet_wrap(~Country)+
theme_light() +
  ggtitle("Average Temperature")


ggplot(pca_data, aes(x=Longitude, y=Latitude,colour=bio12)) +
geom_point(alpha=0.5) +
facet_wrap(~Country)+
theme_light() +
  ggtitle("Average Precipitation")


# Extract specific columns (5 to 23) from pca_data to use for PCA analysis
mat_pca_control <- pca_data[, c(6:24)]

# Perform Principal Component Analysis (PCA) on the selected data
# Center and scale the data before performing PCA
pca_con <- prcomp(mat_pca_control, center = TRUE, scale. = TRUE)

# Visualize the PCA results
# fviz_pca is a function from the factoextra package used for PCA visualization
fviz_pca(pca_con,
col.ind = pca_data$Country,
pointshape = 19,
pointsize = 2,
labelsize = 3,
legend.title = "Environment vs pots",
geom.var = c("arrow", "text"),
col.var = "black",
label="var"
)

# Generate a correlation plot using the corPlot function from the psych package
# This excludes the first column of mat_pca_control for the correlation plot
psych::corPlot((mat_pca_control[,-c(1)]),
               main = 'Correlation chart for all variables')

# Only include chosen uncorrelated variables.
filtered_data <- mat_pca_control[, c(3, 6, 7, 12, 15)]
psych::corPlot(filtered_data, main = 'Unccorelated Variables')

pca_con_filtered <- prcomp(filtered_data, center = TRUE, scale. = TRUE)
fviz_pca(pca_con_filtered,
col.ind = pca_data$Country,# color by groups
pointshape = 19,
pointsize = 2,
labelsize = 3,
legend.title = "Environment vs pots",
geom.var = c("arrow", "text"),
col.var = "black",
label="var"
)

# write uncorrelated variables with no headers for LFMM input.

write.table(filtered_data, "clim.env", sep="\t", quote=F, row.names=F, col.names=F)
```

